---
title: Model diagnostics

---

check_model(Gfap_Fit1)




```{r}
# Add loo for model comparison
Gfap_IntDen_Fit1 <- 
  add_criterion(Gfap_IntDen_Fit1, c("loo", "waic", "bayes_R2"))
```

The models are stored as `.RDS` objects in `Models/Widefield_5x_Ipsilateral_Gfap-Pdgfrb_IntDen` If the file is available in the specified folder, executing the the fit will load the object. For refit, erase the file.

## Model comparison

We perform model comparison by out-of-sample prediction error using the WAIC criteria. By considering both the log-likelihood of the observed data and the effective number of parameters, WAIC provides a balance between model fit and complexity. This approach inherently accounts for the posterior distribution of the parameters rather than relying on specific point estimates. Please note we exclude model 4 (Mdl4) from this comparison because it fulfills another purposes.

Given a data set and a Bayesian model, the WAIC is computed as:

$$
\text{WAIC} = -2 \times \left( \text{LLPD} - p_{\text{WAIC}} \right)
$$

Where: $\text{LLPD}$ is the log point wise predictive density, calculated as the average log likelihood for each observed data point across the posterior samples. $\text{WAIC}$ is the effective number of parameters, computed as the difference between the average of the log likelihoods and the log likelihood of the averages across posterior samples.

We use the `compare_performance` function from the `performance` package (part of the `easystats` environment) [@performance; @easystats; @makowski2019] to display the models comparison parameters, including, WAIC.

First, we visualize the results as a table:

```{r}
#| label: GfapIntDen_Compare
#| include: true
#| warning: false
#| message: false
#| results: false

Gfap_IntDen_Comp <- 
  compare_performance(
    Gfap_IntDen_Fit1, 
    Gfap_IntDen_Fit2, 
    Gfap_IntDen_Fit3, 
    metrics = "all"
    )
Gfap_IntDen_Comp
```

The output depicts that model 3 is the less penalized and has a stronger explained variance (R2 = 0.51). Please note that an estimation for sigma is absent in this model as it was included as a response variable. The preceding offer sufficient grounds to select model 3 (Mdl3) as our best alternative for scientific inference.

We can appreciate the same results using `loo_compare`:

```{r}
#| label: fig-GfapIntDen_Compare
#| include: true
#| warning: false
#| message: false
#| results: false
#| fig-cap: Model camparison by WAIC
#| fig-height: 4
#| fig-width: 5


Gfap_IntDen_W <- 
loo_compare(
  Gfap_IntDen_Fit1, 
  Gfap_IntDen_Fit2, 
  Gfap_IntDen_Fit3,
  criterion = "waic")

# Generate WAIC graph
Gfap_IntDen_WAIC <- 
  Gfap_IntDen_W[, 7:8] %>% 
  data.frame() %>% 
  rownames_to_column(var = "model_name") %>% 
  
ggplot(
  aes(x    = model_name, 
      y    = waic, 
      ymin = waic - se_waic, 
      ymax = waic + se_waic)
  ) +
  geom_pointrange(shape = 21) +
  scale_x_discrete(
    breaks=c("Gfap_IntDen_Fit1", 
             "Gfap_IntDen_Fit2", 
             "Gfap_IntDen_Fit3", 
             "Gfap_IntDen_Fit4"),
    labels=c("Mdl1", 
             "Mdl2", 
             "Mdl3", 
             "Mdl4")
    ) +
  coord_flip() +
  labs(x = "", 
       y = "WAIC (score)",
       title = "") +
  Plot_theme

Gfap_IntDen_WAIC
```

## Model diagnostics

To evaluate sample predictions, we perform the diagnostics for model 3 (Mdl3) using the `pp_check` (posterior predictive checks) function from `brms`. In the graph, 𝘺 shows the data and y\~ the simulated data.

```{r}
#| label: fig-GfapIntDen_Diagnostics
#| include: true
#| warning: false
#| message: false
#| results: false
#| fig-cap: Model diagnostics for GFAP expression
#| fig-height: 4
#| fig-width: 5

set.seed(8807)

# Model 3
Gfap_IntDen_Mdl3_pp <- 
  brms::pp_check(Gfap_IntDen_Fit3, 
                 ndraws = 100) +
  labs(title = "Posterior predictive checks (model 3)",
  subtitle = "Formula: Gfap_IntDen ~ s(DPI, K=5), sigma ~ s(DPI, K=5)") +
  #scale_y_continuous(limits=c(0, 0.15)) +
  Plot_theme  
  
Gfap_IntDen_Mdl3_pp
```

@fig-GfapIntDen_Diagnostics shows that the model predictions (y\~) do not deviate substantially from the data (𝘺). We can use `shinystan` to further explore the model:

```{r}
#| label: GfapIntDen_Shiny
#| include: true
#| warning: false
#| message: false
#| results: false
#| cache: true

#launch_shinystan(Gfap_IntDen_Fit3)
```

# Model results

## Visualizing the results

After validating the model, we plot the full posterior distributions:

```{r}
#| label: fig-GfapIntDen_fit
#| include: true
#| warning: false
#| message: false
#| results: false
#| fig-cap: Posterior for GFAP IntDen
#| fig-width: 5
#| fig-height: 4

set.seed(8807)

# We define the predictor grid
nd <- Gfap_Summary %>% 
  distinct(DPI)

# compute and save the fitted draws
Gfap_Post <- fitted(
  Gfap_IntDen_Fit3, 
  newdata = nd,
  # this line returns the full set of HMC draws
  summary = FALSE) %>% 
  data.frame() %>% 
  set_names(nd %>% pull(DPI)) 
  
Gfap_Post %>% 
  # convert to the long format
  pivot_longer(everything()) %>% 
  mutate(name = fct_relevel(name, "0", "3", "7", "14", "30")) %>% 
  
  ggplot(aes(x = value, y = factor(name))) +
  # adjust the point and interval summary settings as desired
  stat_halfeye(aes(fill = after_stat(level)), .width = c(0.50,0.80,0.90)) +
  scale_fill_brewer() +
  scale_y_discrete(name = "DPI") +
  scale_x_continuous(name = "GFAP (IntDen)",
                     limits = c(0, 1200)) +
  Plot_theme
```

We see an increasing trend up to 14 DPI. We consider it pertinent to further evaluate the causes of the decrease to 30 DPI.

We visualize further the data using the `conditional_effects` function from `brms`. We add graph aesthetics using ggplot.

The `conditional_effects` function display the effects of predictors for brms-fitted models. It computes conditional effects of specified predictors while marginalizing over the remaining predictors. Thereby, it allows to visualization of single variable effects to make the data more interpretable. Mathematically, let $y = f(x_1, x_2, \ldots, x_n)$ denote the modeled relationship between the variable \$ y \$ and the predictor variables \$ x_1, x_2, \ldots, x_n \$. The `conditional_effects` function calculates the conditional expectations $E[y | x_i]$ for a given \$ x_i \$, marginalizing over other predictors. Formally, it can be represented in this manner:

$$
 E[y | x_i] = \int f(x_1, x_2, \ldots, x_n) p(x_1, x_2, \ldots, x_{i-1}, x_{i+1}, \ldots, x_n) dx_1 dx_2 \ldots dx_{i-1} dx_{i+1} \ldots dx_n
$$ In our case, we use the function to visualize the effect of DPI on GFAP IntDen and sigma.

```{r}
#| label: fig-Gfap_IntDen_Sigma_CE
#| include: true
#| warning: false
#| message: false
#| fig-cap: Conditional effects for Gfap_IntDen and (sigma)
#| fig-width: 5
#| fig-height: 4

set.seed(8807)


# Conditional effects for Gfap_IntDen
#####################################

# We convert the estimates to a data frame
Gfap_IntDen_DPI <- 
  conditional_effects(Gfap_IntDen_Fit3)

Gfap_IntDen_DPIdf <- 
  as.data.frame(Gfap_IntDen_DPI$DPI)

Gfap_IntDen_pred <- 
  Gfap_Summary %>%
  data_grid(DPI, n = 10) %>%
  add_predicted_draws(
    Gfap_IntDen_Fit3, 
    ndraws = 50)

Gfap_IntDen_DPI_Fig <-  
  ggplot(
    data  = Gfap_IntDen_DPIdf,
    aes(x = DPI,
        y = estimate__)
    )+
  geom_jitter(
    data  = Gfap_Summary, 
    aes(x = DPI, y=Gfap_IntDen),
    width = 0.5, 
    color = "black", 
    shape = 18, 
    size  = 2.5, 
    alpha = 2/3) +
  geom_ribbon(
    aes(ymin  = lower__, 
        ymax  = upper__),
        alpha = 0.2)+
  geom_line(
    size     = 1,
    position = position_dodge(0.05), 
    color    = "blue") +
  scale_y_continuous(
    name = "GFAP (IntDen)",
    limits = c(0, 1500),
    breaks = seq(0, 1500, 500)
    ) +
  
  scale_x_continuous(name="DPI") +
  Plot_theme

ggsave(
  plot     = Gfap_IntDen_DPI_Fig, 
  filename = "Plots/Widefield_5x_Ipsilateral_Gfap-Pdgfrb_IntDen/Widefield_5x_Ipsilateral_Gfap_IntDen.png", 
  width    = 9, 
  height   = 9, 
  units    = "cm")



# Conditional smooths for sigma
###############################

Gfap_IntDen_Smooths <- 
  conditional_smooths(Gfap_IntDen_Fit3)

Gfap_IntDen_Sigma_df <- 
  as.data.frame(Gfap_IntDen_Smooths$`sigma: s(DPI,k=5)`)

Gfap_IntDen_Sigma_fig <-  
  ggplot(Gfap_IntDen_Sigma_df,
         aes(x = DPI,y = estimate__))+
  geom_ribbon(
    aes(ymin  =lower__, 
        ymax = upper__), 
    alpha = 0.2)+
  geom_line(size  = 1,
            color ="blue",
            position = position_dodge(0.05)) +
  scale_y_continuous(name= expression("GFAP (log)"[sigma])) +
  scale_x_continuous(name="DPI") +
  Plot_theme
          
ggsave(plot     = Gfap_IntDen_Sigma_fig, 
       filename = "Plots/Widefield_5x_Ipsilateral_Gfap-Pdgfrb_IntDen/Widefield_5x_Ipsilateral_Gfap_Sigma_IntDen.png", 
       width    = 9, 
       height   = 9, 
       units    = "cm")

Gfap_IntDen_DPI_Fig | Gfap_IntDen_Sigma_fig
```

@fig-Gfap_IntDen_Sigma_CE shows the splines for GFAP integrated density and sigma. We can see an increasing trend in the integrated density of GFAP until the second week after injury, followed by a slight decrease at 30 DPI. We speculate that the latter could be caused by brain shrinkage. The estimates for sigma show a strong increase in variance at 7 DPI, which is maintained until chronic stages. Please note that the values shown for sigma are not in actual scale, but internally transformed by default in `brms` models.

## Posterior summary

Next, we plot the posterior summary using the `describe_posterior` function from `bayestestR` package [@bayestestR; @makowski2019].

```{r}
#| label: GfapIntDen_DescribePosterior
#| include: true
#| warning: false
#| message: false
#| results: false
#| cache: true

describe_posterior(
  Gfap_IntDen_Fit3,
  effects = "all",
  test = c("p_direction", "rope"),
  component = "all",
  centrality = "median")

modelsummary(Gfap_IntDen_Fit3, 
             shape = term ~ model + statistic,
             centrali2ty = "mean", 
             title = "GFAP integrated density following MCAO",
             statistic = "conf.int",
             gof_omit = 'ELPD|ELDP s.e|LOOIC|LOOIC s.e|WAIC|RMSE',
             output = "Tables/html/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit3_Table.html",
             )

Gfap_IntDen_Fit3_Table <- modelsummary(Gfap_IntDen_Fit3, 
             shape = term ~ model + statistic,
             centrality = "mean", 
             statistic = "conf.int",
             gof_omit = 'ELPD|ELDP s.e|LOOIC|LOOIC s.e|WAIC|RMSE',
             output = "gt")
gt::gtsave (Gfap_IntDen_Fit3_Table, filename = "Tables/tex/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit3_Table.tex")
```

The table displays the regression coefficients. In all cases, the coefficients for splines models are not straightforward to interpret. For this reason, we use the `estimate_relation` and `estimate_smooth` functions from the `modelbased` package [@modelbased] to transform the splines into "linear" regression segments that are easy to interpret.

## Estimate derivative

We estimate the derivative to separate positive from negative trends in the fitted curve and obtain more clear linear fits. In statistical modeling, derivatives provide a measure of how a predictor variable influences the response variable, offering a more nuanced interpretation beyond the sign and magnitude of parameter estimates.

In mathematical terms, the derivative of a smooth function $f(x)$ is defined as:

$$ 
f'(x) = \lim\_{{\Delta x \to 0}} \frac{f(x + \Delta x) - f(x)}{\Delta x} 
$$ {#eq-derivative}

For a Bayesian model where $y = f(x)$, the derivative $\frac{dy}{dx}$ or $f'(x)$ represents the rate of change in the response variable $y$ for a unit change in the predictor variable $x$. If the model is non-linear, this derivative is a function of $x$ itself, allowing us to visualize how this rate of change varies across different values of $x$.

First, we estimate point-to point derivatives:

```{r}
#| label: GfapIntDen_EstimateRelation
#| include: true
#| warning: false
#| message: false
#| results: false
#| cache: true

Gfap_IntDen_est <- 
  estimate_relation(
    Gfap_IntDen_Fit3, 
    at = "DPI", 
    ci = c(0.95))

Gfap_IntDen_est
```

The result show the different "fragments" from the curve with predicted point estimates and their uncertainty. This table is more informative that the raw posterior summary from splines.

To summarize further, we can use `estimate_smooth`function.

```{r}
#| label: GfapIntDen_EstimateSmooth
#| include: true
#| warning: false
#| message: false
#| results: false
#| cache: true
 
Gfap_IntDen_Derivative <- estimate_smooth(Gfap_IntDen_est, x = "DPI")
Gfap_IntDen_Derivative 

gt::gt(round(Gfap_IntDen_Derivative, digits = 2)) %>%
gt::gtsave(filename = "Tables/html/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit3_Deriv.html")

gt::gt(round(Gfap_IntDen_Derivative, digits = 2)) %>%
gt::gtsave(filename = "Tables/tex/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit3_Deriv.tex")
```

Our model predicts a sustained increase (slope = 50.1) of GFAP integrated density up to \~ 16 DPI (Change of 835, 95%CI = 692 - 1223). This is followed by a decrease (slope -36) up to 30 DPI. We do not rule out that this decrease can be mediate by brain shrinkage.

## Visualize derivatives

We visualize the derivatives:

```{r}
#| label: fig-GfapIntDen_EstimateSlopes
#| include: true
#| warning: false
#| message: false
#| results: false
#| fig-cap: Derivatives for GFAP expression
#| fig-height: 4
#| fig-width: 5

Gfap_IntDen_deriv <- 
  estimate_slopes(
    Gfap_IntDen_Fit3,
    trend = "DPI",
     at = "DPI",
    length = 30
    )

# Visualize
Gfap_IntDen_deriv_Fig <- 
  plot(Gfap_IntDen_deriv) + 
  labs(x = "DPI", 
       y = "Effect of DPI",
       title = NULL
       ) +
  Plot_theme +
  theme(legend.position= "none")

ggsave(
  plot     = Gfap_IntDen_deriv_Fig, 
  filename = "Plots/Widefield_5x_Ipsilateral_Gfap-Pdgfrb_IntDen/Widefield_5x_Ipsilateral_Gfap_IntDen_Deriv.png", 
  width    = 9, 
  height   = 9, 
  units    = "cm")

Gfap_IntDen_deriv_Fig 
```

@fig-GfapIntDen_EstimateSlopes shows the pronounced effects of DPI in GFAP expression, with a transition phase around the end of the second week post injury.

# GFAP relation to hemispheric area

Previously, we fitted a forth model (Gfap_IntDen \~ t2(DPI,Tissue_Area)) to investigate the relation between GFAP integrated density and hemispheric area.

## Model diagnostics

We begin by evaluating the sample predictions using `pp_check` for model 4 (Mdl4). In the graph, 𝘺 shows the data and y\~ the simulated data.

```{r}
#| label: fig-GfapIntDen_Hem_Diagnostics
#| include: true
#| warning: false
#| message: false
#| results: false
#| fig-cap: Model diagnostics for GFAP expression / hemispheric area
#| fig-height: 4
#| fig-width: 5

set.seed(8807)

# Model 3
Gfap_IntDen_Mdl4_pp <- 
  brms::pp_check(Gfap_IntDen_Fit4, 
                 ndraws = 100) +
  labs(title = "Posterior predictive checks (model 4)",
  subtitle = "Formula: Gfap_IntDen ~ t2(DPI,Tissue_Area)") +
  #scale_y_continuous(limits=c(0, 0.15)) +
  Plot_theme  
  
Gfap_IntDen_Mdl4_pp
```

We can witness that this model does not imply major deviations from the observed data. Subsequently, we visualize the results using the `conditional_smooths` function from `brms`.

## Visualizaing the results

```{r}
#| label: fig-GfapIntDen_Hemis
#| include: true
#| warning: false
#| message: false
#| fig-cap: GFAP relation with hemispheric area
#| fig-width: 5
#| fig-height: 5

set.seed(8807)

# We convert the estimates to a data frame (uncomment next line to execute)
Gfap_IntDen_Area <-  conditional_smooths(Gfap_IntDen_Fit4)

Gfap_IntDen_Area_Fig  <- 
  plot(Gfap_IntDen_Area, 
       plot = FALSE)[[1]] +
  
  scale_y_continuous(name = expression ("Hemispheric area (mm)"^2)) +
  scale_x_continuous(name="DPI") +
  guides(colour = guide_colourbar(direction = "horizontal")) +
  guides(colour = guide_colourbar(barheight = unit(2, "cm"))) +
  scale_color_gradientn(
    name = "IntDen", 
    colors = viridis(6)) +
  Plot_theme +
  theme(
    legend.position = c(0.17, 0.8))
  
ggsave(
  plot     = Gfap_IntDen_Area_Fig , 
  filename = "Plots/Widefield_5x_Ipsilateral_Gfap-Pdgfrb_IntDen/Widefield_5x_Ipsilateral_Gfap_IntDen-Area.png", 
  width    = 9, 
  height   = 9, 
  units    = "cm")

Gfap_IntDen_Area_Fig
```

@fig-GfapIntDen_Hemis predicts that GFAP integrated intensity will be higher at \~30 DPI with an hemispheric area of 15-20 mm\^2. However, the observations depict that at this time point the hemispheric area is below 10 mm\^2. In this way, the model suggest that the astroglial reactivity is constrained by brain shrinkage.

## Posterior summary

The summarize the posterior:

```{r}
#| label: Gfap_IntDenArea_DescribePosterior
#| include: true
#| warning: false
#| message: false
#| results: false
#| cache: true

describe_posterior(
  Gfap_IntDen_Fit4,
  effects = "all",
  test = c("p_direction", "rope"),
  component = "all",
  centrality = "median")

modelsummary(Gfap_IntDen_Fit4, 
             shape = term ~ model + statistic,
             centrali2ty = "mean", 
             title = "GFAP integrated density following MCAO",
             statistic = "conf.int",
             gof_omit = 'ELPD|ELDP s.e|LOOIC|LOOIC s.e|WAIC|RMSE',
             output = "Tables/html/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit4_Table.html",
             )

Gfap_IntDen_Fit4_Table <- modelsummary(Gfap_IntDen_Fit4, 
             shape = term ~ model + statistic,
             centrality = "mean", 
             statistic = "conf.int",
             gof_omit = 'ELPD|ELDP s.e|LOOIC|LOOIC s.e|WAIC|RMSE',
             output = "gt")
gt::gtsave (Gfap_IntDen_Fit4_Table, filename = "Tables/tex/Widefield_5x_Ipsilateral_Gfap_IntDen_Fit4_Table.tex")
```